---
title: "ACE and Food Insecurity"
subtitle: "Descriptives Tables and Figures"
date: October 22, 2023
author: "Noah Gibson"
format: 
    html: 
      self-contained: true
      theme: spacelab
      toc: true
      number-sections: true
      code-tools: true
      code-fold: true
      df-print: kable
editor: source
execute:
  echo: false
  message: false
  warning: false
editor_options: 
  chunk_output_type: inline
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
library(ggplot2)
library(pilot)
theme_set(theme_pilot())
```

The purpose of this document is to create ~final weighted descriptives and tables. 

To create `ace_fi_tidy_v2.csv`, I merged two PSID datasets: the Childhood Retrospective Circumstances Study (collected in 2014) and the 2015 Individual- and Family-level datasets. `ace_fi_tidy_v2.csv` includes three types of individuals: reference persons, spouses, and long-term partners (see `ace_fi_tidy_v2.qmd` for code). I also downloaded the 2017 and 2019 PSID waves but am still trying to figure out if we need them.

## Download/clean data 

Here is a summary of our variables of interest: 

```{r}
d <- read.csv("ace_fi_tidy_v2.csv")

d <- d |> 
  mutate(race_eth = case_when(race == "White" & spanish_descent == 0 ~ "White",
                              race == "Black" & spanish_descent == 0 ~ "Black",
                              spanish_descent > 0 ~ "Hispanic",
                              (race != "White" & race != "Black" 
                               & spanish_descent == 0) ~ "Other"),
         log_income = log10(income + 1))

d <- d |> 
  filter(relation_2015 == 10) |> 
  select(-ace_fin, -ace_dad_unemp, -ace_mom_unemp, -X, -sex, -wic, -free_lunch,
         -fs_raw, -fs_scaled, -fs_status, -ethnicity, -wealth_w_equity, -relation_2015,
         -avg_fin, -avg_emt, -avg_health, -avg_school, -std_fin, -std_emt, -std_health,
         -std_school, -spanish_descent, -race)

d <- d |> 
  mutate(cum_ace = rowSums(select(d, starts_with("ace_"))))
  
```

```{r}
library(skimr)
skim(d)
```

```{r}
#| label: tidy lca 

d_lca <- d |> 
  mutate(ace_dad_unemp = ace_dad_unemp_0_5 + ace_dad_unemp_6_12 + ace_dad_unemp_13_16,
         ace_mom_unemp = ace_mom_unemp_0_5 + ace_mom_unemp_6_12 + ace_mom_unemp_13_16,
         ace_fin = ace_fin_0_5 + ace_fin_6_12 + ace_fin_13_16,
         ace_welfare = ace_welfare_0_5 + ace_welfare_6_12 + ace_welfare_13_16,
         ace_lonely = ace_lonely_6_12 + ace_lonely_13_16,
         ace_bullied_school = ace_bullied_school_6_12 + ace_bullied_school_13_16,
         ace_bullied_out = ace_bullied_out_6_12 + ace_bullied_out_13_16,
         ace_school_happy = ace_school_happy_6_12 + ace_school_happy_13_16,
         ace_school_safety = ace_school_safety_6_12 + ace_school_safety_13_16)

```

```{r}
#| cache: false
#| results: hide

library(poLCA) 

# Assuming your dataset is named 'ace_data'

d_results <- d_lca |> 
  dplyr::select(starts_with("ace"), id) |>
  dplyr::select(ace_divorce, ace_health, ace_school_health, ace_dad_unemp, 
                ace_mom_unemp, ace_fin, ace_welfare, ace_fin_comp, ace_lonely,
                ace_repeat_school, ace_bullied_school, ace_bullied_out, 
                ace_school_happy, ace_school_safety, id) |> 
  mutate(across(starts_with("ace"), ~ . + 1))

# Assuming your data is stored in the data frame 'your_data'
ace_variables <- colnames(d_results)[grep("^ace", colnames(d_results))]  # Assuming ACE variables start with "ace"

# Create the formula with all ACE variables
formula <- cbind(ace_divorce, ace_health, ace_school_health, ace_dad_unemp, 
                ace_mom_unemp, ace_fin, ace_welfare, ace_fin_comp, ace_lonely,
                ace_repeat_school, ace_bullied_school, ace_bullied_out, 
                ace_school_happy, ace_school_safety)~1


# Run LCA
set.seed(1234)
results6 <- poLCA(formula, data = d_results, nclass = 6, nrep = 25)

posterior_probs <- results6$posterior
class_memberships <- max.col(posterior_probs)

# make sure dataset has same number of dropped observations 
d_0 <- d_lca |> 
  dplyr::select(ace_divorce, ace_health, ace_school_health, ace_dad_unemp, 
                ace_mom_unemp, ace_fin, ace_welfare, ace_fin_comp, ace_lonely,
                ace_repeat_school, ace_bullied_school, ace_bullied_out, 
                ace_school_happy, ace_school_safety, id) |> 
  na.omit()
d_00 <- d |> 
  dplyr::select(-starts_with("ace"))

d_0 <- inner_join(d_0, d_00, by = "id")

# combine with class
d_0 <- cbind(d_0, class_memberships)

```

## Descriptive table by class (ace variables)

Here is a breakdown of responses to ACE questions by class membership. I think this serves as a nice "Table 1" in our paper. 
```{r}
library(gtsummary) # for bold_labels()
library(survey)

d_0 <- d_0 |> 
  mutate(class_memberships = case_when(class_memberships == 1 ~ "Financial",
                                       class_memberships == 2 ~ "High",
                                       class_memberships == 3 ~ "Emotional",
                                       class_memberships == 4 ~ "School-related",
                                       class_memberships == 5 ~ "Limited",
                                       class_memberships == 6 ~ "Parental")) |> 
  mutate(class_memberships = as.factor(class_memberships), 
         class_memberships = relevel(class_memberships, ref = "Limited")) |> 
  mutate(health_status = as.factor(health_status), 
         health_status = relevel(health_status, ref = "poor")) |> 
  mutate(race_eth = as.factor(race_eth), 
         race_eth= relevel(race_eth, ref = "White"))

wgt <- svydesign(id=~id, weights=~weight_2015, nest = TRUE, data = d_0)

wgt |>
  tbl_svysummary(
    include = c(starts_with("ace_"), cum_ace),
    by = class_memberships ,
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{p}%"),
    digits = list(all_categorical() ~ c(1, 1),
                  all_continuous() ~ c(1, 1))
  ) |> 
  modify_header(label = "**Variable**",
                all_stat_cols() ~ "**{level}** ({style_percent(p, digits=1)}%)"
  ) |> 
  modify_caption("Weighted descriptive statistics for all variables (n = 3,927)") |> 
  bold_labels()
```


## Descriptive table by class (controls)

And here are our control variables by class. 
```{r}

wgt |>
  tbl_svysummary(
    include = c(fs, age, income, educ, snap, health_status, female, race_eth),
    by = class_memberships ,
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{p}%"),
    digits = list(all_categorical() ~ c(1, 1),
                  all_continuous() ~ c(1, 1))
  ) |> 
  modify_header(label = "**Variable**",
                all_stat_cols() ~ "**{level}** ({style_percent(p, digits=1)}%)"
  ) |> 
  modify_caption("Weighted descriptive statistics for all variables (n = 3,927)") |> 
  bold_labels()

```

## Model comparisons
As I see it, one key contribution of our paper is improving on cumulative measures of ACEs. So I think it is important to highlight how our approach differs from cumulative measures. This might not be exactly what we want to include in our paper, but I think it is at least interesting to look at. 

Let's compare three models to see which is most statistically informative. Model 1 includes all controls but does not include information on adversity. Model 2 includes all controls and class membership. Model 3 includes all controls and a cumulative measure of adversity. 
```{r}
library(splines)

mod1 <- svyglm(fs ~ age + log_income + female + snap + 
                    race_eth + educ + health_status, design = wgt, family = quasibinomial)

mod2 <- svyglm(fs ~ class_memberships + age + log_income + female + snap + 
                    race_eth + educ + health_status, design = wgt, family = quasibinomial)

mod3 <- svyglm(fs ~ cum_ace + age + log_income + female + snap + 
                    race_eth + educ + health_status, design = wgt, family = quasibinomial)

library(modelsummary)
modelsummary(list(mod1,mod2, mod3), estimate = "{estimate}{stars}")
```

There aren't huge differences between models in terms of BIC and AIC, so we can't really make the case that our approach is improving model fit. This makes senses though, given that a cumulative measure of ACE includes more variation than class membership. 

## Cumulative ACE measure by class
One key way to motivate our paper is by arguing that certain ACE variables are more negatively associated with food insecurity than others. For instance, based on the graph below, previous papers would argue that the `High` class has higher rates of food insecurity than the `School-related` and `Financial` classes. But we find that the `School-related` class has food insecurity rate that is similar to `High`. 

```{r}
library(ggridges)
 
d_0 |> 
  ggplot(aes(y = reorder(class_memberships, cum_ace),
             x = cum_ace, fill = class_memberships))+
  geom_jitter(aes(color = class_memberships),
             position = position_jitter(w = .3, h = .05),
             size = 0.75,
             alpha = 0.15) +
  geom_boxplot(width = .25,
               outlier.shape = NA,
               alpha = 0.6) +
  geom_density_ridges(alpha = .6, scale = .8, 
                      position = position_nudge(y = .175))+
    scale_fill_pilot()+
    scale_color_pilot()+
    theme(legend.position="none")+
    labs(title = "Cumulative ACE Score by Class", y = "Class", 
         x = "Cumulative ACE Score")+
    scale_x_continuous(limits = c(0,24))

```
### Table (for exact numbers when writing)
```{r}

wgt |>
  tbl_svysummary(
    include = cum_ace,
    by = class_memberships ,
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{p}%"),
    digits = list(all_categorical() ~ c(1, 1),
                  all_continuous() ~ c(1, 1))
  ) |> 
  modify_header(label = "**Variable**",
                all_stat_cols() ~ "**{level}** ({style_percent(p, digits=1)}%)"
  ) |> 
  modify_caption("Weighted descriptive statistics for cumulative ACE score (n = 3,927)") |> 
  bold_labels()

```



## Cumulative finacial ACE variables by class

Interestingly, the `School-related` class has low levels of financial adversity as children (I created a cumulative ACE measure just based on the three financial ACE variables). Is it worth including a few graphs like this in our paper?

```{r}
d_0 <- d_0 %>%
  mutate(cum_fin = rowSums(select(d_0, ace_fin, ace_fin_comp, ace_welfare)))

d_0 |> 
  ggplot(aes(y = reorder(class_memberships, cum_fin),
             x = cum_fin, fill = class_memberships))+
  geom_jitter(aes(color = class_memberships),
             position = position_jitter(w = .3, h = .05),
             size = 0.75,
             alpha = 0.15) +
  geom_boxplot(width = .25,
               outlier.shape = NA,
               alpha = 0.6) +
  geom_density_ridges(alpha = .6, scale = .6, 
                      position = position_nudge(y = .175))+
    scale_fill_pilot()+
    scale_color_pilot()+
    theme(legend.position="none")+
    labs(title = "Cumulative Financial ACE Score by Class", y = "Class", 
         x = "Cumulative Financial ACE Score")+
    scale_x_continuous(limits = c(0,8))

```

### Table (for exact numbers when writing)

Weighted mean for cumulative ACE financial variables:
```{r}

# Create a survey design
wgt <- svydesign(id = ~id, weights = ~weight_2015, nest = TRUE, data = d_0)

# Calculate the weighted means by class_membership
weighted_means_by_class <- svyby(~cum_fin, by = ~class_memberships, design = wgt, FUN = svymean)

# Print the weighted means by class_membership
print(weighted_means_by_class)

```


## Cumulative bullying/loneliness ACE variables by class

Now let's take a look at just bullying/loneliness ACE variables. 

```{r}
d_0 <- d_0 %>%
  mutate(cum_bul = rowSums(select(d_0, ace_lonely, ace_bullied_out, ace_bullied_school)))

d_0 |> 
  ggplot(aes(y = reorder(class_memberships, cum_bul),
             x = cum_bul, fill = class_memberships))+
  geom_jitter(aes(color = class_memberships),
             position = position_jitter(w = .3, h = .05),
             size = 0.75,
             alpha = 0.15) +
  geom_boxplot(width = .25,
               outlier.shape = NA,
               alpha = 0.6) +
  geom_density_ridges(alpha = .6, scale = .6, 
                      position = position_nudge(y = .175))+
  scale_fill_pilot()+
  scale_color_pilot()+
  theme(legend.position="none")+
  labs(title = "Cumulative Bullying/Loneliness ACE Score by Class", 
       y = "Class", x = "Cumulative Bullying/Loneliness ACE Score")+
  scale_x_continuous(limits = c(0,7))
```
### Table (for exact numbers when writing)

Weighted mean for cumulative ACE bullying/loneliness variables:
```{r}

# Create a survey design
wgt <- svydesign(id = ~id, weights = ~weight_2015, nest = TRUE, data = d_0)

# Calculate the weighted means by class_membership
weighted_means_by_class <- svyby(~cum_bul, by = ~class_memberships, design = wgt, FUN = svymean)

# Print the weighted means by class_membership
print(weighted_means_by_class)

```



## Food insecurity rates by class

Moving on to our outcome, the `School-related` class has a food insecurity rate that is similar to `High`, which I think is the most interesting finding in our results. (The dashed line is the average food insecurity rate for the sample.)

```{r}

mod2 <- svyglm(fs ~ class_memberships, design = wgt, family = quasibinomial)

library(marginaleffects)

p_class <- predictions(
    mod2,
    type = "response"
    )

p_class <- p_class|> 
  group_by(class_memberships) |> 
  summarise(pred = mean(estimate), lower = mean(conf.low, na.rm = TRUE),
            upper = mean(conf.high, na.rm = TRUE)) 
p_class
```

```{r}
p_class %>%
  ggplot(aes(y=reorder(class_memberships, pred), x = pred))+
    geom_point(alpha = .4, size = 3.5, aes(color = class_memberships))+
    geom_linerange(aes(y=reorder(class_memberships, pred), xmin=lower, xmax=upper,
                        color = class_memberships), 
                    alpha=.9, size=0.75)+
    scale_color_pilot()+
    labs(title = "Food Insecurity Rate by Class", y = "Class",
         x = "Food Insecurity Rate")+
    geom_vline(xintercept = .102, linetype = "dashed", color = "black")+
    theme(legend.position="none")+
    scale_x_continuous(breaks = seq(0, .3, by = 0.05), 
                       labels = format(seq(0, .3, by = 0.05), nsmall = 2))
```

> Why is the food insecurity rate so high for the `School-related` group?

I think we need to try to unpack this a bit in our paper. We know that the `School-related` group had 1.) low levels of financial adversity, and 2.) high levels of emotional adversity. Point 1 leads me to think that we might be able to connect this to literature on coping mechanisms. Is it possible that individuals that struggled financially in childhood have better coping strategies for financial precarity in adulthood? Point 2 makes me think of networks. I know some literature on food insecurity and social networks. Could it be that people in the `School-related` groups have fewer social resources?

### Bootstrapped version 

```{r}

bootstrap_predictions <- function(num_bootstrap_samples, model_formula) {
  # Create an empty matrix to store bootstrap predicted probabilities
  bootstrap_preds <- matrix(NA, nrow = num_bootstrap_samples, 
                            ncol = length(unique(d_0$class_memberships)))
  
  # Bootstrap the predictions
  set.seed(123)  # Set a seed for reproducibility
  
  for (i in 1:num_bootstrap_samples) {
    # Sample with replacement from the dataset
    sampled_data <- d_0[sample(nrow(d_0), replace = TRUE), ]
    
    # Fit the logistic regression model to the resampled data
  boot_model <- svyglm(model_formula, 
                       design = svydesign(id = ~1, weights = ~weight_2015, 
                                          data = sampled_data), 
                       family = quasibinomial)
    
    # Calculate predicted probabilities with weights
    p_class <- predictions(boot_model, type = "response")
    
    # Store the predicted probabilities in the matrix
    bootstrap_preds[i, ] <- tapply(p_class$estimate,
                                   p_class$class_memberships, mean)
  }
  
  # Convert the matrix of bootstrap predictions to a data frame
  bootstrap_preds_df <- as.data.frame(bootstrap_preds)
  
  # Tidy data
  bootstrap_preds_df <- bootstrap_preds_df |> 
  rename( "Limited" = "V1", 
          "Emotional" = "V2",
          "Financial" = "V3", 
          "High" = "V4", 
          "Parental" = "V5", 
          "School-related" = "V6") |> 
  pivot_longer(
    cols = c("Limited":"School-related"),
    names_to = "class_memberships",
    values_to = "pred") |> 
  mutate(class_memberships = as.factor(class_memberships), 
         class_memberships = relevel(class_memberships, ref = "Limited"))
  
  
  return(bootstrap_preds_df)
}

```

```{r}
m1 <- fs ~ class_memberships


d <- bootstrap_predictions(2500, m1)
```

```{r}
mod2 <- svyglm(fs ~ class_memberships, design = wgt, family = quasibinomial)

p_class <- predictions(
    mod2,
    type = "response"
    )

p_class <- p_class|> 
  group_by(class_memberships) |> 
  summarise(pred = mean(estimate), lower = mean(conf.low, na.rm = TRUE),
            upper = mean(conf.high, na.rm = TRUE)) 
```


```{r}
  ggplot(data = p_class,
         aes(y=reorder(class_memberships, pred), x = pred))+
    geom_point(alpha = 1, size = 3.5, 
               aes(y=reorder(class_memberships, pred), x = pred,
                   color = class_memberships))+
    geom_linerange(aes(y=reorder(class_memberships, pred),
                       color = class_memberships, 
                       xmin=lower, xmax=upper), 
                       alpha= 1, size= 1.5)+
    geom_density_ridges(data = d, 
                        alpha = .4, scale = .8,
                        position = position_nudge(y = 0),
                        aes(y = reorder(class_memberships, pred),
                        x = pred, fill = class_memberships))+
    scale_fill_pilot()+
    scale_color_pilot()+
    labs(title = "", y = "Class", 
         x = "Food Insecurity Rate")+
    geom_vline(xintercept = .102, linetype = "dashed", color = "black")+
    theme(legend.position="none")+
    scale_x_continuous(breaks = seq(0, .3, by = 0.05), 
                       labels = format(seq(0, .3, by = 0.05), nsmall = 2))
```


## Predicted probabilities by class

Here is a look at predicted probabilities (which adjust for controls). The `School-related` class has a high predicted probability of food insecurity even when adjusting for key demographic variables in 2015.  

```{r}

mod2 <- svyglm(fs ~ class_memberships + age + log_income + female + snap + 
                    race_eth + educ + health_status, 
               design = wgt, family = quasibinomial)

p_class <- predictions(
    mod2,
    type = "response"
    )

p_class <- p_class|> 
  group_by(class_memberships) |> 
  summarise(pred = mean(estimate), lower = mean(conf.low, na.rm = TRUE),
            upper = mean(conf.high, na.rm = TRUE)) 
p_class
```


```{r}
p_class %>%
  ggplot(aes(y=reorder(class_memberships, pred), x = pred))+
    geom_point(alpha = .4, size = 3.5, aes(color = class_memberships))+
    geom_linerange(aes(y=reorder(class_memberships, pred), 
                       xmin=lower, xmax=upper,
                       color = class_memberships), 
                    alpha=.9, size=0.75)+
    scale_color_pilot()+
    labs(title = "Predicted Probability of Food Insecurity by Class", y = "Class"
       , x = "Pr(Food Insecurity)", 
       subtitle = "Adjusting for variables in model")+
    geom_vline(xintercept = .102, linetype = "dashed", color = "black")+
    theme(legend.position="none")+
    scale_x_continuous(breaks = seq(0, .4, by = 0.05), 
                       labels = format(seq(0, .4, by = 0.05), nsmall = 2))

```

### Bootstrapped version 
```{r}

m2 <- fs ~ class_memberships + age + log_income + female + snap + race_eth + educ + health_status


d2 <- bootstrap_predictions(2500, m2)
```

```{r}

mod2 <- svyglm(fs ~ class_memberships + age + log_income + female + snap + 
                    race_eth + educ + health_status, design = wgt, 
               family = quasibinomial)

p_class <- predictions(
    mod2,
    type = "response"
    )

p_class <- p_class|> 
  group_by(class_memberships) |> 
  summarise(pred = mean(estimate), lower = mean(conf.low, na.rm = TRUE),
            upper = mean(conf.high, na.rm = TRUE)) 
```


```{r}
  ggplot(data = p_class,
         aes(y=reorder(class_memberships, pred), x = pred))+
    geom_point(alpha = 1, size = 3.5, 
               aes(y=reorder(class_memberships, pred), x = pred,
                   color = class_memberships))+
    geom_linerange(aes(y=reorder(class_memberships, pred),
                       color = class_memberships, 
                       xmin=lower, xmax=upper), 
                       alpha= 1, size= 1.5)+
    geom_density_ridges(data = d2, 
                        alpha = .4, scale = .8,
                        position = position_nudge(y = 0),
                        aes(y = reorder(class_memberships, pred),
                        x = pred, fill = class_memberships))+
    scale_fill_pilot()+
    scale_color_pilot()+
    labs(title = "", y = "Class", 
         x = "Pr(Food Insecurity)")+
    geom_vline(xintercept = .102, linetype = "dashed", color = "black")+
    theme(legend.position="none")+
    scale_x_continuous(breaks = seq(0, .3, by = 0.05), 
                       labels = format(seq(0, .3, by = 0.05), nsmall = 2))
```





